---
title: "ARU standardization tutorial"
output: pdf_document
---

## Intro

Data analysis from the following manuscript:

Van Wilgenburg, S. L., Sólymos, P., Kardynal, K. J. and Frey, M. D., 2017. Paired sampling standardizes point count data from humans and acoustic recorders. Avian Conservation and Ecology, 12(1):13. URL: [https://doi.org/10.5751/ACE-00975-120113](https://doi.org/10.5751/ACE-00975-120113)

## Install data package

```{r}
#devtools::install_github("borealbirds/paired")
library(paired)
data(paired)
summary(paired)
x <- paired # make a copy
```

## Preliminaries

Load required packages

```{r}
library(mefa4)
library(MASS)
library(detect)
library(lme4)
```

Relevel survey type: set human point count as the reference

```{r}
x$SurveyType <- relevel(x$SurveyType, "HUM")
```

Remove species that commonly do flyovers or are otherwise inappropriate for pointcounts, set the minimun number of detections

```{r}
(aa <- table(x$SPECIES))
SPP <- names(aa[aa >= 15])
SPP <- SPP[!(SPP %in% c("AMCR","AMGO","BCFR","BLJA","CANG","COLO","COGO","COME","CORA","EVGR","FRGU","GRAJ","UNKN","RESQ","WOSP","WWCR","WISN","SACR","PISI","UNBI","LEYE","GRYE","WOFR","UNKN"))]
x<-x[x$SPECIES %in% SPP,]
nmin <- 15
```

Training and validation data set

```{r}
xv <- droplevels(x[x$RandomSel == 0,])
x <- droplevels(x[x$RandomSel == 1,])
```

## EDR estimation

```{r}
xtdis <- Xtab(Count ~ PKEY + DISTANCE + SPECIES, x, subset=x$SurveyType == "HUM")
for (i in 1:length(xtdis))
    xtdis[[i]] <- as.matrix(xtdis[[i]][,c("0-49 m", "50-100 m", ">100 m")])
ndis <- sapply(xtdis, function(z) sum(rowSums(z)>0))
xtdis <- xtdis[ndis >= nmin]
DDdis <- matrix(c(0.5, 1, Inf), nrow(xtdis[[1]]), 3, byrow=TRUE)
xdis <- nonDuplicated(x, PKEY, TRUE)
xdis <- xdis[rownames(xtdis[[1]]),]

mdis <- list()
for (i in 1:length(xtdis)) {
    m <- cmulti(xtdis[[i]] | DDdis ~ 1, xdis, type = "dis")
    if (!inherits(m, "try-error"))
    mdis[[names(xtdis)[i]]] <- m
}

mdis <- lapply(xtdis, function(z) cmulti(z | DDdis ~ 1, type = "dis"))
edrH <- sapply(mdis, function(z) exp(coef(z)))
names(edrH) <- names(xtdis)
edrH
```

## Singing rate estimation

```{r}
xtdurH <- Xtab(Count ~ PKEY + Interval + SPECIES, x, 
               subset=x$SurveyType == "HUM")
xtdurA <- Xtab(Count ~ PKEY + Interval + SPECIES, x, 
               subset=x$SurveyType == "ARU")
DDdur <- matrix(c(3, 5, 10), nrow(xtdurH[[1]]), 3, byrow=TRUE)
xx <- xdis[rownames(xtdurH[[1]]),]
ndurH <- sapply(xtdurH, function(z) sum(rowSums(z)>0))
ndurA <- sapply(xtdurA, function(z) sum(rowSums(z)>0))

OK <- ndurH >= nmin & ndurA >= nmin
xtdurH <- xtdurH[OK]
xtdurA <- xtdurA[OK]

DDdur2 <- rbind(DDdur, DDdur)
```


Trying to include covariates (e.g. canopy closure)

```{r}
ndurH <- sapply(xtdurH, function(z) sum(rowSums(z)>0))
ndurA <- sapply(xtdurA, function(z) sum(rowSums(z)>0))
maxH <- sapply(xtdurH, function(z) max(rowSums(z)))
maxA <- sapply(xtdurA, function(z) max(rowSums(z)))
OK <- ndurH >= nmin & ndurA >= nmin & maxH > 1 & maxA > 1
xtdurH <- xtdurH[OK]
xtdurA <- xtdurA[OK]
DDdur2 <- rbind(DDdur, DDdur)
```

Store models results for ARU effect on availability

```{r}
mdurA <- list()
mdurH <- list()
mdurHA <- list()
mdurHA1 <- list() 
for (i in 1:length(xtdurA)) {
  yA <- as.matrix(xtdurA[[i]])[,c("0-3 min","3-5 min","5-10 min")]
  yH <- as.matrix(xtdurH[[i]])[,c("0-3 min","3-5 min","5-10 min")]
  mdurA[[names(xtdurA)[i]]] <- cmulti(yA | DDdur ~ 1, type = "rem")
  mdurH[[names(xtdurA)[i]]] <- cmulti(yH | DDdur ~ 1, type = "rem")
  yyy <- rbind(yH, yA)
  aru01 <- rep(0:1, each=nrow(xtdurH[[i]]))
  mdurHA[[names(xtdurA)[i]]] <- cmulti(yyy | DDdur2 ~ 1, type = "rem")
  mdurHA1[[names(xtdurA)[i]]] <- cmulti(yyy | DDdur2 ~ aru01, type = "rem")
}
cfA <- sapply(mdurA, coef)
names(cfA) <- names(mdurA)
cfH <- sapply(mdurH, coef)
names(cfH) <- names(mdurH)
cfHA <- sapply(mdurHA, coef)
names(cfHA) <- names(mdurHA)
phiA <- exp(cfA)
phiH <- exp(cfH)
phiHA <- exp(cfHA) ## Availability from model with ARU effect
```

Confidence Intervals for removal model for availability including a fixed effect for ARU versus human observer

```{r}
ci <- t(sapply(mdurHA1, function(z) confint(z)[2,]))
nrow(ci)

nam <- rownames(ci)[rowSums(is.na(ci)) > 0 | rowSums(sign(ci)) != 0]

ciHuman<-t(sapply(mdurH, function(z) confint(z)[1,]))
ciARU<-t(sapply(mdurA, function(z) confint(z)[1,]))

ciHuman<-round(1-exp(-10*exp(ciHuman)), 3)
nrow(ciHuman)
ciARU<-round(1-exp(-10*exp(ciARU)), 3)
ciARU

p10_H=round(1-exp(-10*exp(cfH)), 3)
p10_A=round(1-exp(-10*exp(cfA)), 3)

availability <- data.frame(ciHuman, p10_H, ciARU,p10_A)
colnames(availability) <- c("LCLHuman", "UCLHuman",
                            "meanHuman","LCLARU","UCLARU","meanARU")
availability <- availability[,c("LCLHuman", "meanHuman",
                                "UCLHuman","LCLARU","meanARU", "UCLARU")] 
```

These are the species where we have:

- reasonable sample size from training data to estimate phi and edr
- phi is not significantly different between aru and human

```{r}
namOK <- rownames(ci)[rowSums(sign(ci)) == 0]
SPP <- intersect(names(edrH), namOK)
SPP <- SPP[!(SPP %in% c("RESQ","UNKN"))]
phiOK <- phiHA[SPP]
edrOK <- edrH[SPP]
```

Validations starts here

```{r}
set.seed(1)
kfold_results <- list()
prop <- 0.75
K <- 10

spp <- "OVEN" # here comes the loop for species

for (k in 1:K) {
    cat("fold", k, "/", K, "\n")
    flush.console()
    calibration_levels <- sample(levels(x$SS), round(prop*nlevels(x$SS)))
    tmp <- x[x$SS %in% calibration_levels,]
    tmp$PKEYm <- droplevels(tmp$PKEYm)
    xt <- as.matrix(Xtab(Count ~ PKEYm + SPECIES, tmp))
    xxx <- nonDuplicated(x, PKEYm, TRUE)
    xxx <- droplevels(xxx[rownames(xt),])

    validation_levels <- levels(x$SS)[!(levels(x$SS) %in%
        calibration_levels)]
    tmpv <- x[x$SS %in% validation_levels,]
    tmpv$PKEYm <- droplevels(tmpv$PKEYm)
    xtv <- as.matrix(Xtab(Count ~ PKEYm + SPECIES, tmpv))
    xxxv <- nonDuplicated(x, PKEYm, TRUE)
    xxxv <- droplevels(xxxv[rownames(xtv),])

    y <- xt[,spp]
    C <- edrOK[spp]^2 * pi * (1-exp(-phiOK[spp]*10))
    off <- rep(log(C), nrow(xxx))

    mod0 <- glmer(y ~ 1 + (1|PKEY), xxx, offset=off, family=poisson)
    mod1 <- glmer(y ~ SurveyType + (1|PKEY), xxx, offset=off, family=poisson)
    delta_estimate <- sqrt(exp(fixef(mod1)[2]))
    aic <- AIC(mod0, mod1)
    aic$dAIC <- aic$AIC - min(aic$AIC)

    yv <- xtv[,spp]
    ## previous offset
    off1 <- rep(log(C), nrow(xxxv))
    mod0v <- glmer(yv ~ 1 + (1|PKEY), xxxv, offset=off1, family=poisson)
    mod1v <- glmer(yv ~ SurveyType + (1|PKEY), xxxv, offset=off1, family=poisson)
    ## offset with delta
    off2 <- ifelse(xxxv$SurveyType == "HUM",
        rep(log(C), nrow(xxxv)), rep(log(C) + log(delta_estimate^2), nrow(xxxv)))
    mod2v <- glmer(yv ~ 1 + (1|PKEY), xxxv, offset=off2, family=poisson)
    mod3v <- glmer(yv ~ SurveyType + (1|PKEY), xxxv, offset=off2, family=poisson)
    aicv <- AIC(mod0v, mod1v, mod2v, mod3v)
    aicv$dAIC <- aicv$AIC - min(aicv$AIC)

    kfold_results[[k]] <- list(
        calibration=aic,
        validation=aicv,
        delta=unname(delta_estimate),
        yc=c(ARU=mean(y[xxx$SurveyType=="ARU"]), HUM=mean(y[xxx$SurveyType=="HUM"])),
        yv=c(ARU=mean(y[xxxv$SurveyType=="ARU"]), HUM=mean(y[xxxv$SurveyType=="HUM"])))
}

sapply(kfold_results, function(z) rank(z$calibration[,"dAIC"]))
sapply(kfold_results, function(z) rank(z$validation[,"dAIC"]))

sapply(kfold_results, function(z) z$calibration[,"dAIC"])
sapply(kfold_results, function(z) z$validation[,"dAIC"])

delta <- sapply(kfold_results, function(z) z$delta)
yratiov <- sapply(kfold_results, function(z) z$yv[1]/z$yv[2])
summary(delta^2)
summary(yratiov)

plot(delta^2, yratiov, xlim=c(0.8,1.1), ylim=c(0.8,1.1), xlab="delta^2",
    ylab="mean(Y_aru) / mean(Y_hum)")
abline(0,1)
abline(h=1,v=1)
```

